\input logicdocument

\heading
Week 4 Problem Set
Your Name
Introduction to Formal Logic
\endheading

Translate the following English sentences into the language of first-order logic.

\problems
\problem{1}
If Ruth is a logician, then she's happy.
	\answer
	$ $
	\endanswer

\problem{2}
Ruth is a logician if she's happy.
	\answer
	$ $
	\endanswer

\problem{3}
Ruth is a logician only if she's happy.
	\answer
	$ $
	\endanswer

\problem{4}
If Ruth and Alonzo are happy, then so is Kurt.
	\answer
	$ $
	\endanswer

\problem{5}
Ruth is happy only if either Alonzo or Kurt is.
	\answer
	$ $
	\endanswer

\problem{6}
If Ruth is a logician, then she'll only be happy if she's working.
	\answer
	$ $
	\endanswer

\problem{7}
Ruth is happy if and only if she's a logician.
	\answer
	$ $
	\endanswer

\problem{8}
If Ruth is happy, then she's not working.
	\answer
	$ $
	\endanswer

\problem{9}
Ruth is working unless she's happy.
	\answer
	$ $
	\endanswer

\problem{10}
Ruth is working unless she's a happy logician.
	\answer
	$ $
	\endanswer

\problem{11}
If either Ruth or Alonzo is working, then neither is happy.
	\answer
	$ $
	\endanswer

\problem{12}
Ruth isn't happy unless she's at work, and in that case, Alonzo is also working.
	\answer
	$ $
	\endanswer

\endproblems

Construct full (not abbreviated) truth tables for the following arguments. For each one, state whether it's valid or invalid.

\problems
\problem{13}
\argument
 A > -B
 -[C ^ B]
\argumentline
 C > -A
\endargument
	\answer
	\truthtable
	 A | B | C : A. > .-.B | -.[.C. ^ .B.] | C. > .-.A
	\truthtableline
	   |   |   :  .   . .  |  . . .   . .  |  .   . . 
	\endtruthtable
	Valid/Invalid?
	\endanswer

\problem{14}
\argument
 -*[[A > B] ^ -A*]
 [-B v A] > B
\argumentline
 A > -[B v A]
\endargument
	\answer
	\truthtable
	 A | B : -.*[[.A. > .B.]. ^ .-.A.*] | [.-.B. v .A.]. > .B | A. > .-.[.B. v .A.]
	\truthtableline
	   |   :  .   . .   . . .   . . .   |  . . .   . . .   .  |  .   . . . .   . 
	\endtruthtable
	Valid/Invalid?
	\endanswer

\endproblems

For the following argument,
\list
a. Use an abbreviated truth table to determine whether it's valid or invalid, given only the connectives. State your decision.

b. If it's invalid, attempt to build a height model that's a counterexample. If the argument is valid, or if no countermodel can be constructed, just say so.

c. State whether the argument is valid or invalid, given both the connectives and the predicates.
\endlist

\problems
\problem{15}
\argument
 -Shorter(ruth,alonzo)
 Shorter(ruth,kurt) > Shorter(ruth,alonzo)
\argumentline
 -[Shorter(alonzo,kurt) v Shorter(ruth,kurt)]
\endargument
	\answerlist
	a.
	\truthtable
	 -.S(r,a) | S(r,k). > .S(r,a) | -.[.S(a,k). v .S(r,k).]
	\truthtableline
	  .       |       .   .       |  . .      .   .
	\endtruthtable
	Valid/Invalid?, given only the connectives.

	b. It's valid.\OR No countermodel is possible.\OR
	\heightmodel
	 
	\endheightmodel

	c. Valid/Invalid?, given the connectives and predicates.
	\endanswerlist

\endproblems

For the following argument, use an abbreviated truth table to determine whether it's valid or invalid. State your decision.

\problems
\problem{16}
\argument
 A v [B > C]
 C > D
 [B > D] > E
\argumentline
 -A > E
\endargument
	\answer
	\truthtable
	 A. v .[.B. > .C.] | C. > .D | [.B. > .D.]. > .E | -.A. > .E
	\truthtableline
	  .   . . .   . .  |  .   .  |  . .   . . .   .  |  . .   .
	\endtruthtable
	Valid/Invalid?
	\endanswer

\endproblems
\bye
