\input logicdocument

\heading
Week 3 Problem Set
Your Name
Introduction to Formal Logic
\endheading

Construct truth tables for the following arguments. For each one, state whether it's valid or invalid.

\problems
\problem{Example}
\argument
 A ^ -B
 C v A
\argumentline
 -[C ^ B]
\endargument
	\answer
	\truthtable
	 A | B | C : A. ^ .-.B | C. v .A | -.[.C. v .B.]
	\truthtableline
	 T | T | T : t.*F .F.t | t.*T .t |*F. .t. T .t
	 T | T | F : t.*F .F.t | f.*T .t |*F. .f. T .t
	 T | F | T : t.*T .T.f | t.*T .t |*F. .t. T .f
	 T | F | F : t.*T .T.f | f.*T .t |*T. .f. F .f
	 F | T | T : f.*F .F.t | t.*T .f |*F. .t. T .t
	 F | T | F : f.*F .F.t | f.*F .f |*F. .f. T .t
	 F | F | T : f.*F .T.f | t.*T .f |*F. .t. T .f
	 F | F | F : f.*F .T.f | f.*F .f |*T. .f. F .f
	\endtruthtable

	Invalid. (You don't need to say why, but it's because row 3 is T T F.)
	\endanswer

\problem{1}
\argument
 -A
 -[B ^ A]
\argumentline
 -B
\endargument
	\answer
	\truthtable
	 A | B : -.A | -.[.B. ^ .A.] | -.B
	\truthtableline
	   |   :  .  |  . . .   . .  |  . 
	\endtruthtable
	Valid/Invalid?
	\endanswer

\problem{2}
\argument
 A v B
 -A v C
 -B v C
\argumentline
 C
\endargument
	\answer
	\truthtable
	 A | B | C : A. v .B | -.A. v .C | -.B. v .C | C
	\truthtableline
	   |   |   :  .   .  |  . .   .  |  . .   .  |  
	\endtruthtable
	Valid/Invalid?
	\endanswer

\problem{3}
\argument
 -*[[A ^ B] v -A*]
 [-B v A] ^ B
\argumentline
 A v -[B v A]
\endargument
	\answer
	\truthtable
	 A | B : -.*[[.A. ^ .B.]. v .-.A.*] | [.-.B. v .A.]. ^ .B | A. v .-.[.B. v .A.]
	\truthtableline
	   |   :  .   . .   . . .   . . .   |  . . .   . . .   .  |  .   . . . .   . 
	\endtruthtable
	Valid/Invalid?
	\endanswer
\endproblems

For each of the following arguments, use an abbreviated truth table to determine whether it's valid or invalid. State your decision.

\problems
\problem{Example}
\argument
 A v -C
 -[A v B]
\argumentline
 -[B v C]
\endargument
	\answer
	\truthtable
	 A. v .-.C | -.[.A. v .B.] | -.[.B. v .C.]
	\truthtableline
	  . T . .  | T. . .   . .  | F. . .   . .
	  .   . .  |  . . . F . .  |  . . . T . .
	  .   . .  |  . .*F.  .*F. |  . . .   . .
	 F.   . .  |  . . .   . .  |  . .F.   . .
	  .   .T.  |  . . .   . .  |  . . .   .*T.
	  .  ..T/*F|  . . .   . .  |  . . .   . .
	\endtruthtable
	Valid. (You don't need to say why, but it's because when we attempted to make both premises T and the conclusion F at the same time, we wound up contradicting ourselves.)
	\endanswer

\problem{4}
\argument
 A v B
 -[C ^ B]
\argumentline
 -C
\endargument
	\answer
	\truthtable
	 A. v .B | -.[.C. ^ .B.] | -.C
	\truthtableline
	  . T .  | T. . .   . .  | F.
	  .   .  |  . . .   . .  |  .
	\endtruthtable
	Valid/Invalid?
	\endanswer

\problem{5}
\argument
 A ^ [B v C]
 -B ^ -[A ^ D]
\argumentline
 -[D v B] ^ C
\endargument
	\answer
	\truthtable
	 A. ^ .[.B. v .C.] | -.B. ^ .-.[.A. ^ .D.] | -.[.D. v .B.]. ^ .C
	\truthtableline
	  . T . . .   . .  |  . . T . . . .   . .  |  . . .   . . . F . 
	  .   . . .   . .  |  . .   . . . .   . .  |  . . .   . . .   . 
	\endtruthtable
	Valid/Invalid?
	\endanswer

\endproblems

For each of the following arguments,
\list
a. Use an abbreviated truth table to determine whether it's valid or invalid, given only the connectives. State your decision.

b. If it's invalid, attempt to build a height model that's a counterexample. If the argument is valid, or if no countermodel can be constructed, just say so.

c. State whether the argument is valid or invalid, given both the connectives and the predicates.
\endlist

\problems
\problem{Example}
\argument
 Taller(ruth,alonzo) v Taller(ruth,kurt)
 -[Shorter(kurt,alonzo) ^ Taller(ruth,kurt)]
\argumentline
 Taller(ruth,kurt) v Shorter(kurt,alonzo)
\endargument
	\answerlist
	a.
	\truthtable
	 T(r,a). v .T(r,k) | -.[.S(k,a). ^ .T(r,a).] | T(r,k). v .S(k,a)
	\truthtableline
	       . T .       | T. .      .   .      .  |       . F .
	       .   .       |  . .      . F .      .  |   *F  .   .  *F
	       .   .   F   |  . . F    .   .      .  |       .   .
	   *T  .   .       |  . .      .   .      .  |       .   .
	       .   .       |  . .      .   .   T  .  |       .   .
	\endtruthtable
	Invalid, given only the connectives.

	b.
	\heightmodel
	 Alonzo < Ruth = Kurt
	\endheightmodel
	\OR
	\heightmodel
	 Alonzo < Ruth < Kurt
	\endheightmodel

.	(You only need to give one countermodel, but I've given two for illustration.)

	c. Invalid, given the connectives and predicates.

.	(You don't need to say why, but it's because the height model depicts a situation in which $T(r,a)$ is T, $T(r,k)$ is F, and $S(k,a)$ is F, and we know from the abbreviated truth table that this makes both premises T and the conclusion F.)

	\endanswerlist

\problem{6}
\argument
 Shorter(ruth,kurt) v Taller(ruth,alonzo)
 -[Taller(alonzo,kurt) ^ Taller(ruth,alonzo)]
\argumentline
 -Taller(alonzo,kurt)
\endargument
	\answerlist
	a.
	\truthtable
	 S(r,k). v .T(r,a) | -.[.T(a,k). ^ .T(r,a).] | -.T(a,k)
	\truthtableline
	       .   .       |  . .      .   .      .  |  .
	\endtruthtable
	Valid/Invalid?, given only the connectives.

	b. It's valid.\OR No countermodel is possible.\OR
	\heightmodel
	 
	\endheightmodel

	c. Valid/Invalid?, given the connectives and predicates.
	\endanswerlist

\problem{7}
\argument
 Taller(ruth,alonzo) ^ Taller(alonzo,kurt)
 -Taller(ruth,alonzo) v Taller(alonzo,kurt)
\argumentline
 Taller(ruth,kurt) v -Taller(alonzo,kurt) 
\endargument
	\answerlist
	a.
	\truthtable
	 T(r,a). ^ .T(a,k) | -.T(r,a). v .T(a,k) | T(r,k). v .-.T(a,k)
	\truthtableline
	       .   .       |  .      .   .       |       .   . .
	\endtruthtable
	Valid/Invalid?, given the connectives.

	b. It's valid.\OR No countermodel is possible.\OR
	\heightmodel
	 
	\endheightmodel

	c. Valid/Invalid?, given the connectives and predicates.
	\endanswerlist

\endproblems
\bye
